Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-26 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision on final part of section 2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2900 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f14c174edd245f1dbd8e13c5b368822026a633b8

parameter constant. Additionally, they developed a combined hetereoscedastic version 

@@

addititve effects conditional on linear covariates are not intuitive

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression 
trees where binary splits are made based on an angular distance measure employed to evalute node homogeneity.
%which is evaluated between 
%the observed response values and the conditional mean direction of the circular response variable
%and in that way serves as a measure of homogeneity within subgroups. 
%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%%subset of observations.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not only for 
linear but also for a circular covariate, the tree only considers effects in the mean direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.

@@

In already existing generalized linear or additive approaches to circular regression modeling 
usually the location parameter $\mu$ of is regressed on 
covariates in the general form 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} x).
\end{equation}
with a circular intercept $\mu_0$ and $x$ being a vector of covariates
with the corresponding vector of coefficients $\beta$ that are to be estimated.
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an interval of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
The concentration parameter $\kappa$ is mostly estimated as a constant, however,
\cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on covariates
employing the exponential function as a link function while $\mu$ is assumed to be a fixed
value. But regardless of whether $\mu$ or $\kappa$ depend on covariates,
both variations have shown to be very complex in their optimization procedure and hence
challenging in their application. Consequently, these difficulties intensify even more in 
a combined model where the sets of regression 
coefficients for the two distribution parameters $\mu$ and $\kappa$ respectively are reestimated 
alternatingly in an iterative optimization algorithm. Hence, regardless of difficulties in practice, 
the presented theory provides a first approach to probabilistic regression modeling of circular data, 
however, the parameters are not regressed simultaneously on covariates as for example in a more general
case by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
true underlying effect is an abrupt shift.
One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
is to employ tree-based models as presented in the following Section.
%% FIXME: better description of difficulties?

>>>>

parameter constant. Additionally, they developed a combined heteroscedastic version 

@@

additive effects conditional on linear covariates are not intuitive

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression 
trees where binary splits are made based on an angular distance measure employed to evaluate node homogeneity.
%which is evaluated between 
%the observed response values and the conditional mean direction of the circular response variable
%and in that way serves as a measure of homogeneity within subgroups. 
%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%%subset of observations.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not only for 
linear but also for a circular covariate, the tree only considers effects in the mean direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.

@@

In already existing generalized linear or additive approaches to circular regression modeling 
usually the location parameter $\mu$ is regressed on covariates in the general form 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} z).
\end{equation}
with a circular intercept $\mu_0$, a vector of covariates $z$
with the corresponding vector of coefficients $\beta$ that are to be estimated,
and a link function $g$.
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an interval of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
constant value. Additionally to these two model variations where only one of the two distribution 
parameters depends on covariates a combination of them is introduced in a joint model. There, the 
two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
in alternation within an iterative optimization algorithm. 
Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
more general case by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
However, in practice the incorporated optimization procedure has shown to be very complex, already
for only one parameter being regressed on covariates and consequently even more so in the combined model.
This is mainly due to characteristics of the log-likelihood function of the employed von Mises distribution
with the plugged-in link function. In particular, the function is not globally concave and possibly very 
sharp local maxima can prevent the algorithm from converging.
Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
true underlying effect is an abrupt shift.
%% FIXME: difficulties in modeling smooth effects on the unit circle
One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
is to employ tree-based models as presented in the following Section.
